Competing Bandits: \\ The Perils of Exploration under Competition

### 2019-04-03 paper/ec19-rebuttal.txt

Apologies for brevity! We'll discuss all issues in revision.

Several important differences compared to [MSW] [26], as discussed on page3:

In [MSW], users don't observe anything about firms' actual performance, and instead choose based on complicated Bayesian reasoning. We consider a much more realistic model with "reputation signals". This model is amenable to numerical simulations (whereas simulating the priors in [MSW] is computationally very difficult). Moreover, we can separate data vs. reputational advantage (Section5).

[MSW] changes "competitiveness" via behavioral model ("HardMax" -> "HardMaxWithRandomness" -> "SoftMax"). We focus on different "channel": early -> simultaneous -> late entry (and investigate the "first-mover advantage").

[MSW] is theory-only, with strong but "asymptotic" results ("after sufficiently many rounds"). Our findings are more quantitative: whether effects are sufficiently strong in realistic time.

(Finding11 is a good example: as we emphasize right after, the effect is not observed until fairly large T. Put differently, no effect when/if such T is unrealistic, even though this effect is predicted by [MSW].)

Thus, theorems from [MSW] have no direct bearing on our simulations.

Re lack of proofs:

We believe theorems wouldn't not add much value. Indeed, we are in realm of  "asymptotic" theorems about stylized Economic models, which provide some "hard" intuition but no direct bearing on the world. But [MSW] already has an elaborate set of such results, with similar high-level conclusions. So, we'd need to go beyond the first-order conclusions to add value.

While amenable to simulations, our model appears difficult to analyze:

    - intricate feedback loop (performance -> reputations -> users -> performance),

    - mean reputation, most connected to our intuition, is sometimes bad predictor in competition (Sections 3,6),

    - mathematical tools from regret-minimization only produce "asymptotic" results.

We proved *something*: if incumbent starts so early that it has permanent "first-order stochastic dominance" over entrant as standalone algorithm, then it has some advantage under competition. This result is quite weak, yet not easy to prove, and we didn't see much added value. Perhaps we should have included it, just to illustrate the difficulty of analysis.

Re significance:

General adage, we don't know which intuitions/effects are significant until we test. In particular, our intuitions are all about mean reputations, which is sometimes wrong thing to track (Sections 3,6). Also, some effects matter only after very long time (ex: Finding 11).

Welfare implications are at least somewhat surprising.

We didn't have much/any intuition about "data advantage" vs. "reputational advantage"

Re monopoly [Review3]: No findings here, just assumption: monopolist chooses
Dynamic Greedy (DG) because it is easier to deploy in practice (page 2). Implicitly, users have no "outside option": the service provided is an improvement over not having it. This is plausible with free ad-supported platforms such as Yelp or Google. Pricing-for-quality modifies firms' utilities in all scenarios, not just under monopoly. Also a valid model, worth investigating.

[Review3] Re changing competitiveness: "channels" from literature (Lerner Index [3] or Herfindahl-Hirschman Index) just aren't applicable,  we and [MSW] invent channels that are (more details don't fit here).

[Review3] Lower "memory size"  induces noise which we cannot easily characterize. Such noise is also unrealistic, as reputation signals usually average over many datapoints. We remove "noise" in main experiments (but study it in Section7).

[Review1] Agreed, "adaptive strategies" (ex: DG then TS) are interesting! However, "fixed strategies" are arguably more realistic: detecting competition, committing to new algorithms, and adapting/deploying them takes very long time in practice (background: https://arxiv.org/abs/1606.03966). Much larger strategy space, obviously. We did some pilot experiments, where we found that changing the algorithm did not matter.

### 2019-04-03 paper/ec19-rebuttal.txt

Apologies for brevity! We'll discuss all issues in revision.

Several important differences compared to [MSW] [26], as discussed on page3:

In [MSW], users don't observe anything about firms' actual performance, and instead choose based on complicated Bayesian reasoning. We consider a much more realistic model with "reputation signals". This model is amenable to numerical simulations (whereas simulating the posteriors in [MSW] is computationally very difficult). Moreover, we can separate data vs. reputational advantage (Section5).

@@

    - mathematical tools from regret-minimization only produce "asymptotic" results.

We proved *something*: if incumbent starts so early that it has permanent "first-order stochastic dominance" over entrant as standalone algorithm, then it has some advantage under competition. This result is quite weak, yet not easy to prove, and we didn't see much added value. Perhaps we should have included it, just to illustrate the difficulty of analysis.

Re significance:

General adage, we don't know which intuitions/effects are significant until we test. In particular, much of our intuitions are about mean reputations, which is sometimes wrong thing to track (Sections 3,6). Also, some effects matter only after very long time (see Finding 11).

@@

We didn't have much/any intuition about "data advantage" vs. "reputational advantage"

Re monopoly [Review3]: No findings here, just assumption: monopolist chooses
Dynamic Greedy (DG) because it is easier to deploy in practice (page 2). Implicitly, users have no "outside option": the service provided is an improvement over not having it. This is plausible with free ad-supported platforms such as Yelp or Google. Pricing-for-quality modifies firms' utilities in all scenarios, not just under monopoly. Also a valid model, worth investigating.

[Review3] Re changing competitiveness: "channels" from literature (Lerner Index [3] or Herfindahl-Hirschman Index) just aren't applicable,  we and [MSW] invent channels that are (more details don't fit here).

[Review3] Lower "memory size"  induces noise that we cannot easily characterize. Such noise is also unrealistic, as reputation signals usually average over many datapoints. We remove "noise" in main experiments (but study it in Section7).

[Review1] Agreed, "adaptive strategies" are interesting! However, "fixed strategies" are arguably more realistic: detecting competition, committing to new algorithms, and adapting/deploying them takes very long time in practice (background: https://arxiv.org/abs/1606.03966). Much larger strategy space, obviously. We did some pilot experiments, where we found that changing the algorithm did not matter.

### 2019-04-03 paper/ec19-rebuttal.txt

Welfare implications are at least somewhat surprising.

We didn't have much/any intuition about "data advantage" vs. "reputational advantage"

Re monopoly [Review3]: No findings here, just assumption: monopolist chooses
Dynamic Greedy (DG) because it is easier to deploy in practice (page 2). Implicitly, users have no "outside option": the service provided is an improvement over not having it. This is plausible with free ad-supported platforms such as Yelp or Google. Pricing-for-quality modifies firms' utilities in all scenarios, not just under monopoly. Also a valid model, worth investigating.

[Review3] Re changing competitiveness: "channels" from literature (Lerner Index [3] or Herfindahl-Hirschman Index) just aren't applicable in our setting, whereas we and [MSW] identify channels that are. (More details don't fit here).

@@

[Review3] Lower "memory size"  induces noise that we cannot easily characterize. Such noise is also unrealistic, as reputation signals usually average over many datapoints. We remove "noise" in main experiments (but study it in Section7).

[Review1] Agreed, "dynamic strategies" (ex: DG then TS) are interesting! However, "fixed strategies" are arguably more realistic: detecting competition, committing to new algorithms, and adapting/deploying them takes very long time in practice (background: https://arxiv.org/abs/1606.03966). Much larger strategy space, obviously. We did some pilot experiments, where we found that changing the algorithm did not matter.

### 2019-04-29 paper/content/conclusion.tex
Some post-rebuttal edits. Also, removed old color-edits.

\asedit{We consider a stylized duopoly setting where firms simultaneously learn from and compete for users. We showed that competition may not always induce firms to commit to better exploration algorithms, resulting in welfare losses for consumers. The primary reason is that exploration may have short-term reputational consequences that lead to more naive algorithms winning in a long-term competition. Allowing one firm to have a head start, a.k.a. the first-mover advantage, incentivizes the first-mover to deploy ``better" algorithms, which in turn leads to better welfare for consumers. Finally, we isolate the component of the first-mover advantage that is due to having more initial data, and find that even a small amount of this ``data advantage" leads to substantial long-term market power.}

>>>>

We consider a stylized duopoly setting where firms simultaneously learn from and compete for users. We showed that competition may not always induce firms to commit to better exploration algorithms, resulting in welfare losses for consumers. The primary reason is that exploration may have short-term reputational consequences that lead to more naive algorithms winning in a long-term competition. Allowing one firm to have a head start, a.k.a. the first-mover advantage, incentivizes the first-mover to deploy ``better" algorithms, which in turn leads to better welfare for consumers. Finally, we isolate the component of the first-mover advantage that is due to having more initial data, and find that even a small amount of this ``data advantage" leads to substantial long-term market power.

### 2019-04-29 paper/content/introduction.tex
Some post-rebuttal edits. Also, removed old color-edits.

Online platforms routinely deploy A/B tests, and are increasingly adopting  more sophisticated exploration methodologies based on \emph{multi-armed bandits}, a well-known framework for making decisions under uncertainty and trading off exploration and exploitation (making good near-term decisions). While deploying ``better" learning algorithms for exploration would improve performance, this is not necessarily beneficial under competition, even putting aside the deployment/maintenance costs. In particular, excessive experimentation may hurt a platform's reputation and decrease its market share in the near term. This would leave the learning algorithm with less users to learn from, which may further degrade the platform's performance relative to competitors who keep learning and improving from \emph{their} users, and so forth. \asedit{Taken to the extreme, such dynamics may create a ``death spiral" effect when the vast majority of customers eventually switch to competitors.}

@@

We consider a \emph{permanent duopoly} in which both firms start at the same time, as well as \emph{temporary monopoly}: a duopoly with a first-mover. Accordingly, the intensity of competition in the model varies from ``permanent monopoly" (just one firm) to ``incumbent" (the first-mover in temporary monopoly) to permanent duopoly to ``entrant" (late-arriver in temporary monopoly).

@@

\xhdr{Main findings.}
We find that in the permanent duopoly, competition incentivizes firms to choose the ``greedy algorithm", and even more so if the firm is a late arriver in a market. This algorithm also prevails under monopoly, simply because it tends to be easier to deploy. Whereas the incumbent in the temporary monopoly is incentivized to deploy a more advanced exploration algorithm. As a result, consumer welfare is highest under temporary monopoly. \asedit{We find strong evidence of the ``death spiral" effect mentioned earlier; this effect is strongest under permanent duopoly.}

@@

\xhdr{Additional findings.}
We investigate the ``first-mover advantage" phenomenon in more detail. Being first in the market gives free data to learn from (a ``data advantage") as well as a more definite, and possibly better reputation compared to an entrant (a ``reputation advantage"). We run additional experiments so as to isolate and compare these two effects. \asedit{We find that either effect alone leads to a significant advantage under competition. The data advantage is larger than reputation advantage when the incumbent commits to a more advanced bandit algorithm.}

\asedit{Data advantage is significant from the anti-trust perspective, as a possible barrier to entry. We find that even a small amount ``data advantage" gets amplified under competition, causing a large difference in eventual market shares. This observation runs contrary to prior work  \cite{varian2018artificial,lambrecht2015can,bajari2018impact}, which studied learning without competition, and found that small amounts of additional data do not provide significant improvement in eventual outcomes. We conclude that competition dynamics -- that firms compete as they learn over time -- are pertinent to these anti-trust considerations.}

@@

While reputation scores already introduce some noise and randomness into users' choices, we consider an alternative choice rule in which the noise/randomness is explicit: a small fraction of users choose a firm uniformly at random. We confirm the theoretical intuition that better algorithms prevail if the parameter -- \asedit{essentially, the expected number ``random users"} -- is sufficiently large. However, we find that this effect is negligible for some smaller but still ``relevant" parameter values.

\xhdr{Discussion.}
The present paper is an experimental counterpart to \cite{CompetingBandits-itcs18}, which considered a similar duopoly model and obtained a number of theoretical results with ``asymptotic" flavor. For the sake of analytical tractability, that paper makes a somewhat unrealistic simplification that users do not observe any signals about firms' ongoing performance (instead, users choose between firms according to the firms' Bayesian-expected rewards). The high-level conclusion in \cite{CompetingBandits-itcs18} is an inverted-U relationship similar to ours, but the strength of competition is varied in a different way, using assumptions about (ir)rational user behavior.

@@

@@

Whether data gives competitive advantage
has been studied theoretically in \cite{varian2018artificial,
  lambrecht2015can} and empirically in
\cite{bajari2018impact}.
While these papers find that small amounts
  of additional data do not provide significant improvement,
 % in training a more predictive model,
  they focus on learning in isolation.
  %, which overlooks certain aspects in a competition.
%\swdelete{These studies consider the effect of additional data in terms of helping towards solving a machine learning problem in isolation and argue that it does not play a large role.}
\asdelete{We ask how additional data helps when firms simultaneously solve a machine learning problem and compete for consumers. Due to the competition dynamics in our model, we find that even a small amount of additional data tends to help a lot.}
The first-mover advantage has been well-studied in other settings in economics and marketing, see survey \cite{kerin1992first}.

>>>>

Online platforms routinely deploy A/B tests, and are increasingly adopting  more sophisticated exploration methodologies based on \emph{multi-armed bandits}, a well-known framework for making decisions under uncertainty and trading off exploration and exploitation (making good near-term decisions). While deploying ``better" learning algorithms for exploration would improve performance, this is not necessarily beneficial under competition, even putting aside the deployment/maintenance costs. In particular, excessive experimentation may hurt a platform's reputation and decrease its market share in the near term. This would leave the learning algorithm with less users to learn from, which may further degrade the platform's performance relative to competitors who keep learning and improving from \emph{their} users, and so forth. Taken to the extreme, such dynamics may create a ``death spiral" effect when the vast majority of customers eventually switch to competitors.

@@

We consider a \emph{permanent duopoly} in which both firms start at the same time, as well as \emph{temporary monopoly}: a duopoly with a first-mover. Accordingly, the intensity of competition in the model varies from ``permanent monopoly" (just one firm) to ``incumbent" (the first-mover in temporary monopoly) to permanent duopoly to ``entrant" (late-arriver in temporary monopoly).%
\footnote{\asedit{We consider the ``permanent monopoly" scenario for comparison only, without presenting any findings. We just assume that a monopolist chooses
the greedy algorithm, because it is easier to deploy in practice. Implicitly, users have no ``outside option": the service provided is an improvement over not having it (and therefore the monopolist is not incentivized to deploy better learning algorithms). This is plausible with free ad-supported platforms such as Yelp or Google.}}

@@

\xhdr{Main findings.}
We find that in the permanent duopoly, competition incentivizes firms to choose the ``greedy algorithm", and even more so if the firm is a late arriver in a market. This algorithm also prevails under monopoly, simply because it tends to be easier to deploy. Whereas the incumbent in the temporary monopoly is incentivized to deploy a more advanced exploration algorithm. As a result, consumer welfare is highest under temporary monopoly. We find strong evidence of the ``death spiral" effect mentioned earlier; this effect is strongest under permanent duopoly.

@@

\xhdr{Additional findings.}
We investigate the ``first-mover advantage" phenomenon in more detail. Being first in the market gives free data to learn from (a ``data advantage") as well as a more definite, and possibly better reputation compared to an entrant (a ``reputation advantage"). We run additional experiments so as to isolate and compare these two effects. We find that either effect alone leads to a significant advantage under competition. The data advantage is larger than reputation advantage when the incumbent commits to a more advanced bandit algorithm.

Data advantage is significant from the anti-trust perspective, as a possible barrier to entry. We find that even a small amount ``data advantage" gets amplified under competition, causing a large difference in eventual market shares. This observation runs contrary to prior work  \cite{varian2018artificial,lambrecht2015can,bajari2018impact}, which studied learning without competition, and found that small amounts of additional data do not provide significant improvement in eventual outcomes. We conclude that competition dynamics -- that firms compete as they learn over time -- are pertinent to these anti-trust considerations.

@@

We also consider an alternative choice rule with explicit noise/randomness: a small fraction of users choose a firm uniformly at random.%
\footnote{\asedit{Reputation scores already introduce some noise into users' choices. However, the amount of noise due to this channel is typically small, both in our simulations and in practice, because reputation signals average over many datapoints.}}

 We confirm the theoretical intuition that better algorithms prevail if the parameter -- essentially, the expected number ``random users" -- is sufficiently large. However, we find that this effect is negligible for some smaller but still ``relevant" parameter values.

\subsection{Discussion}

The present paper is an experimental counterpart to \cite{CompetingBandits-itcs18}, which considered a similar duopoly model and obtained a number of theoretical results with ``asymptotic" flavor. For the sake of analytical tractability, that paper makes a somewhat unrealistic simplification that users do not observe any signals about firms' ongoing performance. Instead, users choose between firms according to the firms' Bayesian-expected rewards. The strength of competition is varied in a different way, using assumptions about (ir)rational user behavior. \asedit{For these reasons, the theorems from \cite{CompetingBandits-itcs18} have no direct bearing on our simulations.} However, their high-level conclusion in is an inverted-U relationship similar to ours.

@@

\asedit{While amenable to simulations, our model appears difficult to analyze. This is for several reasons:
%
intricate feedback loop from performance to reputations to users to performance;
%
mean reputation, most connected to our intuition, is sometimes bad predictor in competition (see Sections~\ref{sec:isolation} and~\ref{sec:revisited});
%
mathematical tools from regret-minimization would only produce ``asymptotic" results, which does not seem to suffice.
We believe that resolving first-order theoretical questions would not add much value to this paper, because of \cite{CompetingBandits-itcs18}. We are in realm of ``asymptotic" theorems about stylized Economic models, which provide some mathematical intuition but no direct bearing on the world. But \cite{CompetingBandits-itcs18} already has an elaborate set of such results, with similar high-level conclusions. So, one would need to go beyond the first-order conclusions to add value.}

@@

Whether data gives competitive advantage
has been studied theoretically in \cite{varian2018artificial,
  lambrecht2015can} and empirically in
\cite{bajari2018impact}.
While these papers find that small amounts
  of additional data do not provide significant improvement,
 % in training a more predictive model,
  they focus on learning in isolation.
  %, which overlooks certain aspects in a competition.
%\swdelete{These studies consider the effect of additional data in terms of helping towards solving a machine learning problem in isolation and argue that it does not play a large role.}
The first-mover advantage has been well-studied in other settings in economics and marketing, see survey \cite{kerin1992first}.

### 2019-04-29 paper/content/model.tex
Some post-rebuttal edits. Also, removed old color-edits.

Before the game starts, each firm commits to an MAB algorithm, and uses this algorithm to choose its actions. Each algorithm receives a ``warm start": additional $T_0$ agents that arrive before the game starts, and interact with the firm as described above. \asedit{The warm start ensures that each firm has a meaningful reputation when competition starts.} Each firm's objective is to maximize its market share: the fraction of users who chose this firm.

@@

\xhdr{Agents.}
\asedit{Firms compete on a single dimension, quality of service, as expressed by agents' rewards.}
\OMIT{\footnote{One can think that there are two quality types - a reward of 1 is a high type and a reward of 0 is a low type}}
Agents are myopic and non-strategic: they \asedit{would like} to choose among the firms so as to maximize their expected reward \gaedit{(i.e. select the firm with the highest quality)}, without attempting to influence the firms' learning algorithms or rewards of the future users. Agents are not well-informed: they only receive a rough signal about each firm's performance before they choose a firm, and no other information.

@@

\asedit{We are mainly interested in qualitative differences between the three classes.} For concreteness, we fix one algorithm from each class. Our pilot experiments indicate that our findings do not change substantially if other algorithms are chosen. For technical reasons, we consider Bayesian versions initialized with a ``fake" prior (\ie not based on actual knowledge). We consider:

>>>>

Before the game starts, each firm commits to an MAB algorithm, and uses this algorithm to choose its actions. Each algorithm receives a ``warm start": additional $T_0$ agents that arrive before the game starts, and interact with the firm as described above. The warm start ensures that each firm has a meaningful reputation when competition starts. Each firm's objective is to maximize its market share: the fraction of users who chose this firm.

@@

\xhdr{Agents.}
Firms compete on a single dimension, quality of service, as expressed by agents' rewards. 
\OMIT{\footnote{One can think that there are two quality types - a reward of 1 is a high type and a reward of 0 is a low type}}
Agents are myopic and non-strategic: they would like to choose among the firms so as to maximize their expected reward (i.e. select the firm with the highest quality), without attempting to influence the firms' learning algorithms or rewards of the future users. Agents are not well-informed: they only receive a rough signal about each firm's performance before they choose a firm, and no other information.

@@

We are mainly interested in qualitative differences between the three classes. For concreteness, we fix one algorithm from each class. Our pilot experiments indicate that our findings do not change substantially if other algorithms are chosen. For technical reasons, we consider Bayesian versions initialized with a ``fake" prior (\ie not based on actual knowledge). We consider:

### 2019-04-29 paper/content/perf_in_iso.tex
Some post-rebuttal edits. Also, removed old color-edits.

\asedit{We also use Figure~\ref{prelim_means} to choose a reasonable time-horizon for the subsequent experiments, as $T=2000$. The idea is, we want $T$ to be large enough so that algorithms performance starts to plateau, but small enough such that algorithms are still learning.}

>>>>

We also use Figure~\ref{prelim_means} to choose a reasonable time-horizon for the subsequent experiments, as $T=2000$. The idea is, we want $T$ to be large enough so that algorithms performance starts to plateau, but small enough such that algorithms are still learning.
